A Standard for the Scholarly Citation of Archaeological Data

### 2017-09-14 analysis/paper/paper.Rmd
tweaking some figures

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how to these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviours and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). Archaeologists can be found all over this spectrum because of the diversity of archaeological research, however, our view is that archaeology is for the most part a restricted data-sharing field. There are few standardised and widely used methods for sharing and citing data, a challenge for archaeological data sharing that we hope this paper will help to address.

@@

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2017), we include the entire R code used for all the analysis and visualizations contained in this paper in our SOM at http://dx.doi.org/10.17605/OSF.IO/KSRUZ. Also in this version-controlled compendium are the raw data for all the tests reported here. All of the figures, tables and statistical test results presented here can be independently reproduced with the code and data in this repository. In our SOM our code is released under the MIT licence, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2017).

@@

Our sample of authors that we wrote email to focused on Journal of Archaeological Science articles on experiments with stone artefacts that were published during 2009 and 2015. We focused on stone artefact experiments because the all research materials in these projects are generated by the researcher, so there are no other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymised so the specific authors we contacted are not identifiable.

@@

```{r sharing-rates}
tbl <- readr::read_csv("../figures/tbl_studies_of_data_sharing.csv")
knitr::kable(tbl, caption = "Summary of a selection of previous studies of data sharing in various fields.")
```

@@

```{r, fig.cap = "Relationships between data centres and file formats. The ten most frequently found file formats are shown here, in the data centres with these files from the top ten data centres by number of DOIs. Open file formats for structured data, such as CSV, are not common."}
library(tidyr)

@@

plot_structured_data_files_per_file_tally <- 
ggplot(structured_data_files_per_file_tally,
       aes(str_wrap(datacentre, 20), 
           format)) +
  geom_point(aes(size = n), 
             shape = 16, 
             #alpha = 0.4,
             colour = viridis(3)[3]) +
  geom_text(aes(label = prettyNum(n, big.mark = ",")), size = 3) +
  scale_size_continuous(range=c(2,25), guide = FALSE) +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1)) +
  coord_flip() +
  xlab("Data centre") +
  ylab("File format")

ggsave("../figures/plot_structured_data_files_per_file_tally.png",
       plot_structured_data_files_per_file_tally,
       width = 10)

knitr::include_graphics("plot_structured_data_files_per_file_tally.png")
```

@@

# Ethics and data sharing

@@

The status of research data is slowly changing in many fields from a privately-held commodity traded for co-authorship and other status tokens to part of the public commons, accessible to anyone with an internet connection. This change in status has been fastest in 'big science' fields, such as astronomy, physics, and oceanography, that use expensive instruments that generate data shared with large teams of collaborators to produce new discoveries (Reichman and Uhlir 2003). However, in 'smaller' sciences such as archaeology, where collaborative groups are small, data sharing is still mostly ad hoc and disaggregated. Our pilot studies support this with an overall sharing rate of `r sharing_rate`% in responses to private requests for data, and `r data_are_avail_perc`% of sampled journal articles with openly available data, in a wide variety of formats and locations. In exmaning archaeological data deposited in repositories, we found that most of the contents are PDF files of reports, and only a very small fraction of the data files are structured data such as spreadsheets. Archaeologists still have a long way to go in making their data easier to reuse by sharing data in plain text structured data formats such as CSV. We found that legislation requiring data sharing resulted in a substantial increase in the rate of data deposition, and that archaeologists are sensitive to fees charged by a respository.

@@

@@

@@

@@

@@

@@

>>>>

The surveys above show that data sharing practices tend to be a private activity, but where public sharing does occur, what forms does it take, and how to these vary across different fields of research?  Wouters and Haak (2017) surveyed 1,200 researchers about data sharing behaviours and found a spectrum from ‘intensive data-sharing fields’ to ‘restricted data-sharing fields’. Intensive data-sharing fields often depend on access to global databases, for example human genetics, while restricted data-sharing fields depend mostly on data collected by the individual researcher or their immediate collaborators, such as digital humanities. These differences in disciplinary norms have implications for data ownership (intensive: collectively owned, willing to share; restrictive: owned by the individual, reluctant to share) and the effort required to share (intensive: low effort because data are prepared for sharing as an integral part of the research process, restrictive: high effort because data are prepared for sharing after the research, or apart from it, in response to a request from another researcher). Archaeologists can be found all over this spectrum because of the diversity of archaeological research, however, our view is that archaeology is, for the most part, currently a restricted data-sharing field. There are few standardised and widely used methods for sharing and citing data, a challenge for archaeological data sharing that we hope this paper will help to address.

@@

To enable re-use of our materials and improve reproducibility and transparency according to the principles outlined in Marwick (2017), we include the entire R code used for all the analysis and visualizations contained in this paper in our compendium at http://dx.doi.org/10.17605/OSF.IO/KSRUZ. Also in this version-controlled compendium are the raw data for all the tests reported here. All of the figures and quantitative results presented here can be independently reproduced with the code and data in this repository. In our compendium our code is released under the MIT licence, our data as CC-0, and our figures as CC-BY, to enable maximum re-use (for more details, see Marwick 2017).

@@

Our sample of authors that we wrote email to focused on _Journal of Archaeological Science_ articles on experiments with stone artefacts that were published during 2009 and 2015. We focused on stone artefact experiments because the all research materials in these projects are generated by the researcher, so there are no other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymised so the specific authors we contacted are not identifiable.

@@

```{r sharing-rates}
tbl <- readr::read_csv("../figures/tbl_studies_of_data_sharing.csv")
knitr::kable(tbl, caption = "Summary of a selection of previous studies of data sharing in various fields, from Spencer (2010). In the study by Peng et al., the authors did not attempt to contact other researchers, but noted the availability of data as described in the published article.")
```

@@

```{r}
library(tidyr)

@@

library(ggrepel)
plot_structured_data_files_per_file_tally <- 
ggplot(structured_data_files_per_file_tally,
       aes(str_wrap(datacentre, 20), 
           format)) +
      geom_point(aes(size = n), 
             shape = 16, 
             #alpha = 0.4,
             colour = viridis(3)[3]) +
    geom_text_repel(aes(label = prettyNum(n, big.mark = ",")), 
                  size = 2.5,
                  colour = "black", 
                  segment.colour="black",
                  min.segment.length = unit(0.5, "lines"), 
                  nudge_x = 0.3, 
                  nudge_y = 0.5) +
  scale_size_continuous(range=c(2,25), guide = FALSE) +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1)) +
  coord_flip() +
  xlab("Data centre") +
  ylab("File format")

ggsave("../figures/plot_structured_data_files_per_file_tally.png",
       plot_structured_data_files_per_file_tally,
       width = 10)
```

```{r, fig.cap = "Relationships between data centres and file formats. The ten most frequently found file formats are shown here, in the data centres with these files from the top ten data centres by number of DOIs. Open file formats for structured data, such as CSV, are not common."}
knitr::include_graphics("../figures/plot_structured_data_files_per_file_tally.png")
```

@@

# What are the ethics of data sharing?

@@

The status of research data is slowly changing in many fields from a privately-held commodity traded for co-authorship and other status tokens to part of the public commons, accessible to anyone with an internet connection. This change in status has been fastest in 'big science' fields, such as astronomy, physics, and oceanography, that use expensive instruments that generate data shared with large teams of collaborators to produce new discoveries (Reichman and Uhlir 2003). However, in 'smaller' sciences such as archaeology, where collaborative groups are small, data sharing is still mostly ad hoc and disaggregated. Our pilot studies, while small and limited in many ways, support this with an overall sharing rate of `r sharing_rate`% in responses to private requests for data, and `r data_are_avail_perc`% of sampled journal articles with openly available data, in a wide variety of formats and locations. In exmanining archaeological data deposited in repositories, we found that most of the contents are PDF files of reports, and only a very small fraction of the data files are structured data such as spreadsheets. Archaeologists still have a long way to go in making their data easier to reuse by sharing data in plain text structured data formats such as CSV. We found that legislation requiring data sharing resulted in a substantial increase in the rate of data deposition, and that archaeologists are sensitive to fees charged by a respository.

@@

Kyzas PA, Loizou KT, and Ioannidis JP. (2005) Selective reporting biases in cancer prognostic factor studies. Journal of the National Cancer Institute 97, 1043-1055.

@@

McCullough BD, McGeary KA, Harrison TD. (2006) Lessons from the JMCB Archive. Journal of Money, Credit, and Banking 38(4), 1093-1107. doi:10.1353/mcb.2006.0061

McCullough BD and Vinod HD. (2005) Comment: Econometrics and Software. Journal of Economic Perspectives 17(1), 223-224.

@@

Peng RD, Dominici F, and Zeger SL. (2006) Reproducible Epidemiological Research. American Journal of Epidemology 163, 783-789 doi: 10.1093/aje/kwj093

@@

Reidpath DD, and Allotey PA. Data Sharing in Medical Research: An Empirical Investigation. Bioethics 15(2), 125- 134 (2002). doi: 10.1111/1467-8519.00220

@@

Savage CJ, Vickers AJ (2009) Empirical Study of Data Sharing by Authors Publishing in PLoS Journals. PLoS ONE 4(9): e7078. doi:10.1371/journal.pone.0007078

@@

Spencer, Hilary (2010). Thoughts on the sharing of data and research
materials and the role of journal policies. Retrieved September 7, 2017, from https://web.stanford.edu/~vcs/Nov21/hilary_spencer_rdcscsJan2010.pdf
